\pkg{DNest4}: Diffusive Nested Sampling in \proglang{C++} and \proglang{Python}

### 2016-05-13 paper/ms.tex
random edits

%% an abstract and keywords
\Abstract{In probabilistic (Bayesian) inferences, we typically want to compute
properties of the posterior distribution, describing knowledge of
unknown quantities in the context of a particular dataset and the assumed
prior information. The marginal likelihood (also known as the ``evidence'')
is a key quantity in Bayesian model averaging. The Diffusive Nested Sampling algorithm, a variant of Nested Sampling, is a powerful tool for generating
posterior samples and estimating marginal likelihoods. It is effective at
solving complex problems including many where the posterior distribution is
multimodal or has strong dependencies between variables. The key requirement
is that the likelihood function be fast to evaluate.
\pkg{DNest4} is an open source,
multi-threaded implementation of this algorithm in
modern \proglang{C++} (specifically, \proglang{C++11}),
along with associated utilities including: i)
a \pkg{Python} package allowing basic use without \proglang{C++} coding;
ii) \pkg{RJObject}, a class template for finite mixture models;
iii) Experimental support for models implemented in \proglang{Julia}; and
iv) An experimental \proglang{Python} tool to
generate \pkg{C++} code from a model specification.
In this paper we demonstrate \pkg{DNest4} usage through examples including
simple Bayesian data analysis, finite mixture models, and Approximate
Bayesian Computation (ABC).
}

@@

When discussing computational matters, the prior is
often written $\pi(\theta)$, the likelihood $L(\theta)$,
and the marginal likelihood $Z$. A popular alternative name for the marginal
likelihood, which emphasizes its role in Bayesian model averaging,
is the ``evidence''.

@@

A defining feature of NS is that it works with a sequence of
{\em constrained prior} distributions, proportional to $\pi$ but
restricted to regions of the parameter space where $L(\theta)$
is above some threshold $\ell$:
\begin{align}
p(\theta; \ell) &=
\frac{\pi(\theta)\mathds{1}\left[L(\theta) > \ell\right]}{X(\ell)}
\label{eqn:constrained_prior}
\end{align}
where
\begin{align}
X(\ell) &= \int \pi(\theta) \mathds{1}\left[L(\theta) > \ell\right] \, d\theta
\end{align}
is the amount of prior mass which has likelihood greater than $\ell$.
In the standard NS framework, the sequence of $\ell$ values is selected
so that $X(\ell)$ shrinks by a factor $\approx e^{-1/N}$ per iteration, where
$N$ is the number of particles used.
Sampling from the constrained priors (Equation~\ref{eqn:constrained_prior})
is required, and Markov Chain Monte Carlo (MCMC) is a popular method of doing
this, although alternatives exist \citep[e.g.][]{multinest}.

@@

\section{Dependencies and installation}
\pkg{DNest4} can be obtained from the \pkg{git} repository located at
\begin{center}
{\tt https://github.com/eggplantbren/DNest4/}\\
\end{center}
and is licensed under the
MIT license. To obtain and compile \pkg{DNest4},
you require \pkg{git} and a recent version of \pkg{g++}. The \proglang{Python}
packages \pkg{NumPy}, \pkg{matplotlib},
\pkg{Pandas}, and \pkg{Cython} are also needed.
To download and compile \pkg{DNest4}
the following steps are sufficient:

@@

\section*{Acknowledgements}
It is a pleasure to thank Ewan Cameron (Oxford), Dan
Foreman-Mackey (Washington), David Hogg (NYU), Daniela Huppenkothen (NYU),
Scott Sisson (UNSW), and Yanan Fan (NYU) for valuable discussions.

>>>>

%% an abstract and keywords
\Abstract{In probabilistic (Bayesian) inferences, we typically want to compute
properties of the posterior distribution, describing knowledge of
unknown quantities in the context of a particular dataset and the assumed
prior information. The marginal likelihood (also known as the ``evidence'')
is a key quantity in Bayesian model averaging. The Diffusive Nested Sampling algorithm, a variant of Nested Sampling, is a powerful tool for generating
posterior samples and estimating marginal likelihoods. It is effective at
solving complex problems including many where the posterior distribution is
multimodal or has strong dependencies between variables. The key requirement
is that the likelihood function be fast to evaluate.
\pkg{DNest4} is an open source (MIT licensed),
multi-threaded implementation of this algorithm in
modern \proglang{C++} (specifically, \proglang{C++11}),
along with associated utilities including: i)
a \pkg{Python} package allowing basic use without \proglang{C++} coding;
ii) \pkg{RJObject}, a class template for finite mixture models;
iii) Experimental support for models implemented in \proglang{Julia}; and
iv) An experimental \proglang{Python} tool to
generate \pkg{C++} code from a model specification.
In this paper we demonstrate \pkg{DNest4} usage through examples including
simple Bayesian data analysis, finite mixture models, and Approximate
Bayesian Computation (ABC).
}

@@

When discussing computational matters, the prior distribution for parameters
is
often written $\pi(\theta)$, the likelihood $L(\theta)$,
and the marginal likelihood $Z$. A popular alternative name for the marginal
likelihood, which emphasizes its role in Bayesian model averaging,
is the ``evidence''.

@@

A defining feature of NS is that it works with a sequence of
{\em constrained prior} distributions, proportional to $\pi$ but
restricted to regions of the parameter space where $L(\theta)$
is above some threshold $\ell$:
\begin{align}
p(\theta; \ell) &=
\frac{\pi(\theta)\mathds{1}\left[L(\theta) > \ell\right]}{X(\ell)}
\label{eqn:constrained_prior}
\end{align}
where
\begin{align}
X(\ell) &= \int \pi(\theta) \mathds{1}\left[L(\theta) > \ell\right] \, d\theta
\end{align}
is the amount of prior mass which has likelihood greater than $\ell$.
In the standard NS framework, the sequence of $\ell$ values is selected
so that $X(\ell)$ shrinks by a factor $\approx e^{-1/N}$ per iteration, where
$N$ is the number of particles used.

Sampling from the constrained priors (Equation~\ref{eqn:constrained_prior})
is required, and Markov Chain Monte Carlo (MCMC) is a popular method of doing
this, although alternatives exist \citep[e.g.][]{multinest}.

@@

\section{Dependencies and installation}
The following instructions apply to Unix-like operating systems such as
GNU/Linux, Mac OS X, and FreeBSD.

\pkg{DNest4} can be obtained from the \pkg{git} repository located at
\begin{center}
{\tt https://github.com/eggplantbren/DNest4/}\\
\end{center}
and is licensed under the
MIT license. To obtain and compile \pkg{DNest4},
you require \pkg{git} and a recent version of \pkg{g++}. The \proglang{Python}
packages \pkg{NumPy}, \pkg{matplotlib},
\pkg{Pandas}, and \pkg{Cython} are also needed.
To download and compile \pkg{DNest4}
the following steps are sufficient:

@@

\section*{Acknowledgements}
It is a pleasure to thank Anna Pancoast (Harvard),
Ewan Cameron (Oxford), David Hogg (NYU), and Daniela Huppenkothen (NYU)
for valuable discussions.
